High-order Solution Transfer between Curved Triangular Meshes

### 2024-09-07 doc/paper.tex
Final updates for revised manuscript. Addressing a few more reviewer comments. Adding \ref commands for (optionally) highlighting the changes.

@@

Solution transfer between meshes is a common problem that
has practical applications to many methods in
computational physics. For example, by allowing the underyling
computational domain to change during a simulation, computational
effort can be focused dynamically to resolve relevant features
of a numerical solution. This so-called mesh adaptivity typically
requires translating the numerical solution from the old mesh to the new,
i.e. solution transfer \cite{Babuska1978,Peraire1987,Pain2001,Iske2004}. As another example, Lagrangian or particle-based
methods treat each node in the mesh as a particle and so with each timestep the
mesh travels \emph{with} the fluid \cite{MR3023731}.
However, over (typically short) time the mesh
becomes distorted and suffers a loss in element quality which eventually causes
inaccurate solutions. To overcome this, the
domain must be remeshed or rezoned and the solution must be
transferred (remapped) onto the new mesh configuration. A related more general
class of methods is the Arbitrary Lagrangian-Eulerian (ALE) method, which
are not constrained by a fixed computational mesh (the Eulerian approach) or
by a fixed fluid flow (the Lagrangian approach) and typically combine the benefits
of both approaches \cite{Hirt1974,MR3302403}.

@@

The problem of conservative interpolation
has been considered previously for straight sided meshes.
However, both to allow for greater
geometric flexibility and for high order of convergence, this paper
addresses the case of curved isoparametric meshes. This is based on the
recent interest in so-called high-order methods \cite{Wang2013}, which have the ability to
produce highly accurate solutions with low dissipation and low dispersion error.
However, these methods typically require curved meshes to obtain the high accuracy.

@@

Conservative solution transfer has been around since the advent of ALE,
and as a result much of the existing literature focuses on mesh-mesh
pairs that will occur during an ALE-based simulation. Many ALE methods
modify the mesh during simulation: when flow-based
mesh distortion occurs, elements are typically ``flipped'' (e.g. a
diagonal is switched in a pair of elements) or elements are subdivided
or combined. These operations are inherently local, hence the solution
transfer can be done locally across known neighbors. Typically, this
locality is crucial to solution transfer methods. In \cite{Margolin2003},
the transfer is based on partitioning elements of the updated mesh into
components of elements from the old mesh and ``swept regions'' from
neighbouring elements. In \cite{Kucharik2008}, the (locally) changing
connectivity of the mesh is addressed. In \cite{Garimella2007}, the
local transfer is done on polyhedral meshes.

@@

In addition to conservation, there are many other considerations for a
solution transfer method that are important in various applications. For
example, since our focus is on high-order methods, these typically produce
oscillations in the solutions that can be non-physical and break the numerical
solvers. This can happen for the standard Galerkin projections that we
consider here, which means that e.g. discontinuities such as shocks in the
solution could become nonphysical after our solution transfer. Various
solutions to this have been proposed before, often referred to as limiting or
bound/monotonicity preserving methods, see e.g. \cite{MR3302403,Anderson2018}
where several approaches are studied including a high-order extension of the
flux-corrected transport (FCT) method. Other considerations for a solution
transfer scheme include the preservation of additional constraints, such
as the divergence or the curl of the solution.

In this paper, an algorithm for conservative solution transfer between curved
meshes will be described. This method applies to meshes in \(\reals^2\).
Application to meshes in \(\reals^3\) is a direction for future research,
though the geometric kernels (see Chapter~\ref{sec:bezier-intersection})
become significantly more challenging to describe and implement. In
addition, the method will assume that every element in the
target mesh is contained in the donor mesh. This ensures that the solution
transfer is \emph{interpolative}. In the case where all target elements are
partially covered, \emph{extrapolation} could be used to extend a solution
outside the domain, but for totally uncovered elements there is no clear
correspondence to elements in the donor mesh.

@@

A \emph{B\'{e}zier curve} is a mapping from the unit interval
that is determined by a set of control points
\(\left\{\bm{p}_j\right\}_{j = 0}^n \subset \reals^d\).
For a parameter \(s \in \left[0, 1\right]\), there is a corresponding
point on the curve:
\begin{equation}
b(s) = \sum_{j = 0}^n \binom{n}{j} (1 - s)^{n - j} s^j \bm{p}_j \in
  \reals^d.
\end{equation}
This is a combination of the control points weighted by
each Bernstein basis function
\(B_{j, n}(s) = \binom{n}{j} (1 - s)^{n - j} s^j\).
Due to the binomial expansion
\(1 = (s + (1 - s))^n = \sum_{j = 0}^n B_{j, n}(s)\),
a Bernstein basis function is in
\(\left[0, 1\right]\) when \(s\) is as well. Due to this fact, the
curve must be contained in the convex hull of its control points.

@@

A \emph{valid} B\'{e}zier triangle is one which is
diffeomorphic to \(\utri\), i.e. \(b(s, t)\) is bijective and has
an invertible Jacobian everywhere. We must also have the orientation
preserved, i.e. the Jacobian must have positive determinant. For example, in
Figure~\ref{fig:inverted-element}, the image of \(\utri\) under
the map \(b(s, t) = \left[\begin{array}{c c} (1 - s - t)^2 + s^2 & s^2 + t^2
\end{array}\right]^T\) is not valid because the Jacobian is zero along
the curve \(s^2 - st - t^2 - s + t = 0\) (the dashed line). Elements that
are not valid are called \emph{inverted} because they have regions with
``negative area''. For the example, the image \(b\left(\utri\right)\)
leaves the boundary determined by the edge curves: \(b(r, 0)\),
\(b(1 - r, r)\) and \(b(0, 1 - r)\) when \(r \in \left[0, 1\right]\).
This region outside the boundary is traced twice, once with
a positive Jacobian and once with a negative Jacobian.
\begin{figure}
  \includegraphics{inverted_element.pdf}
  \centering
  \captionsetup{width=.75\linewidth}
  \caption{The B\'{e}zier triangle given by \(b(s, t) = \left[
    (1 - s - t)^2 + s^2 \; \; s^2 + t^2 \right]^T\) produces an
    inverted element. It traces the same region twice, once with
    a positive Jacobian (the middle column) and once with a negative
    Jacobian (the right column).}
  \label{fig:inverted-element}
\end{figure}

@@

Consider a donor mesh \(\mathcal{M}_D\) with shape function basis
\(\phi_D^{(j)}\) and a known field \(\bm{q}_D = \sum_j d_j \phi_D^{(j)}\)
\footnote{This is somewhat a simplification. In CG (continuous Galerkin, as
opposed to DG or discontinuous Galerkin), some coefficients will be paired
with multiple shape functions, such as the coefficient at a vertex node.}
and a
target mesh \(\mathcal{M}_T\) with shape function basis \(\phi_T^{(j)}\)
(Figure~\ref{fig:donor-target-pair}).
Each shape function \(\phi\) corresponds to a
given isoparametric curved element (see Section~\ref{sec:curved-elements})
\(\mathcal{T}\) in one of these meshes and has
\(\operatorname{supp}(\phi) = \mathcal{T}\).
Additionally, the shape functions are polynomial degree \(p\) (see
Section~\ref{subsec:shape-functions} for a discussion of shape functions),
but the degree of the donor mesh need not be the same as that of the
target mesh. We assume that both meshes cover the same domain \(\Omega \subset
\reals^2\), however we really only require the donor mesh to cover the
target mesh.

We seek the \(L_2\)-optimal interpolant \(\bm{q}_T = \sum_j t_j \phi_T^{(j)}\):
\begin{equation}
\left \lVert \bm{q}_T - \bm{q}_D \right \rVert_2 =
\min_{\bm{q} \in \mathcal{V}_T}
\left \lVert \bm{q} - \bm{q}_D \right \rVert_2
\end{equation}
where \(\mathcal{V}_T = \operatorname{Span}_j\left\{\phi_T^{(j)}\right\}\)
is the function space defined on the target mesh. Since this is optimal in
the \(L_2\) sense, by differentiating with respect to each \(t_j\) in
\(\bm{q}_T\) we find the weak form:
\begin{equation}
\int_{\Omega} \bm{q}_D \phi_T^{(j)} \, dV =
  \int_{\Omega} \bm{q}_T \phi_T^{(j)} \, dV, \qquad \text{for all } j.
\end{equation}
If the constant function \(1\) is contained in \(\mathcal{V}_T\),
conservation follows from the weak form and linearity of the integral
\begin{equation}
\int_{\Omega} \bm{q}_D \, dV =
  \int_{\Omega} \bm{q}_T \, dV.
\end{equation}
Expanding \(\bm{q}_D\) and \(\bm{q}_T\) with respect to their coefficients
\(\bm{d}\) and \(\bm{t}\), the weak form gives rise to a linear system
\begin{equation}\label{eq:weak-form-system}
M_T \bm{t} = M_{TD} \bm{d}.
\end{equation}
Here \(M_T\) is the mass matrix for
\(\mathcal{M}_T\) given by
\begin{equation}
\left(M_T\right)_{ij} = \int_{\Omega} \phi_T^{(i)} \phi_T^{(j)} \, dV.
\end{equation}
In the discontinuous Galerkin case, \(M_T\) is block diagonal with blocks
that correspond to each element, so \eqref{eq:weak-form-system} can be
solved locally on each element \(\mathcal{T}\) in the target mesh. By
construction, \(M_T\) is symmetric and sparse since \(\left(M_T\right)_{ij}\)
will be \(0\) unless \(\phi_T^{(i)}\) and \(\phi_T^{(j)}\) are supported
on the same element \(\mathcal{T}\). In the continuous case, \(M_T\) is
globally coupled since coefficients corresponding to boundary nodes interact
with multiple elements. The matrix \(M_{TD}\) is a ``mixed'' mass matrix
between the target and donor meshes:
\begin{equation}
\left(M_{TD}\right)_{ij} = \int_{\Omega} \phi_T^{(i)} \phi_D^{(j)} \, dV.
\end{equation}
Boundary conditions can be imposed on the system by fixing some coefficients,
but that is equivalent to removing some of the basis functions which may
in turn make the projection non-conservative. This is because the removed
basis functions may have been used in \(1 = \sum_j u_j \phi_{T}^{(j)}\).

@@

For special cases, e.g. ALE methods, the initial brute-force search
can be reduced to \(\bigO{1}\). This could be enabled by tracking
the remeshing process so a correspondence already exists. If a full
mapping from donor to target mesh exists, the process of computing
\(M_T\) and \(M_{TD} \bm{d}\) can be fully parallelized across
elements of the target mesh, or even across shape functions
\(\phi_T^{(j)}\). Spatial binning techniques for mesh elements could also be
used to improve the initial brute-force search as in \cite{Weiss16_imr}.

@@

Convergence results are shown in Figure~\ref{fig:composite-errors}
and confirm the expected orders. Since \(\zeta_1\) is a cubic polynomial,
one might expect the solution transfer on the cubic mesh to be \emph{exact}.
This expectation would hold in the superparametric case (i.e. where the
elements are straight sided but the shape functions are higher degree).
However, when the elements aren't straight sided the isoparametric mapping
changes the underlying function space.

@@

>>>>

%\newcommand{\rev}[1]{\textcolor{red}{#1}}
\newcommand{\rev}[1]{#1}

@@

Solution transfer between meshes is a common problem that
has practical applications to many methods in
computational physics. \rev{For example, by allowing the underyling
computational domain to change during a simulation, computational
effort can be focused dynamically to resolve relevant features
of a numerical solution. This so-called mesh adaptivity typically
requires translating the numerical solution from the old mesh to the new,
\rev{i.e. solution transfer \cite{Babuska1978,Peraire1987,Pain2001,Iske2004}}}. As another example, Lagrangian or particle-based
methods treat each node in the mesh as a particle and so with each timestep the
\rev{mesh travels \emph{with} the fluid \cite{MR3023731}.
However, over (typically short) time the mesh
becomes distorted and suffers a loss in element quality which eventually causes
inaccurate solutions.} To overcome this, the
domain must be remeshed or rezoned and the solution must be
transferred (remapped) onto the new mesh configuration. \rev{A related more general
class of methods is the Arbitrary Lagrangian-Eulerian (ALE) method, which
are not constrained by a fixed computational mesh (the Eulerian approach) or
by a fixed fluid flow (the Lagrangian approach) and typically combine the benefits
of both approaches \cite{Hirt1974,MR3302403}.}

@@

The problem of conservative interpolation
has been considered previously for straight sided meshes.
However, both to allow for greater
geometric flexibility and for high order of convergence, this paper
addresses the case of curved isoparametric meshes. \rev{This is based on the
recent interest in so-called high-order methods \cite{Wang2013}, which have the ability to
produce highly accurate solutions with low dissipation and low dispersion error.
However, these methods typically require curved meshes to obtain the high accuracy.}

@@

Conservative solution transfer has been around since the advent of ALE,
and as a result much of the existing literature focuses on mesh-mesh
pairs that will occur during an ALE-based simulation. \rev{Many ALE methods
modify the mesh during simulation: when flow-based
mesh distortion occurs, elements are typically ``flipped'' (e.g. a
diagonal is switched in a pair of elements) or elements are subdivided
or combined.} These operations are inherently local, hence the solution
transfer can be done locally across known neighbors. Typically, this
locality is crucial to solution transfer methods. In \cite{Margolin2003},
the transfer is based on partitioning elements of the updated mesh into
components of elements from the old mesh and ``swept regions'' from
neighbouring elements. In \cite{Kucharik2008}, the (locally) changing
connectivity of the mesh is addressed. In \cite{Garimella2007}, the
local transfer is done on polyhedral meshes.

@@

\rev{In addition to conservation, there are many other considerations for a
solution transfer method that are important in various applications. For
example, since our focus is on high-order methods, these typically produce
oscillations in the solutions that can be non-physical and break the numerical
solvers. This can happen for the standard Galerkin projections that we
consider here, which means that e.g. discontinuities such as shocks in the
solution could become nonphysical after our solution transfer. Various
solutions to this have been proposed before, often referred to as limiting or
bound/monotonicity preserving methods, see e.g. \cite{MR3302403,Anderson2018}
where several approaches are studied including a high-order extension of the
flux-corrected transport (FCT) method. Other considerations for a solution
transfer scheme include the preservation of additional constraints, such
as the divergence or the curl of the solution.}

In this paper, an algorithm for conservative solution transfer between curved
meshes will be described. This method applies to meshes in \(\reals^2\).
Application to meshes in \(\reals^3\) is a direction for future research,
though the geometric kernels (see Chapter~\ref{sec:bezier-intersection})
become significantly more challenging to describe and implement. In
addition, the method will assume that every element in the
target mesh is contained in the donor mesh. This ensures that the solution
transfer is \rev{\emph{interpolative}}. In the case where all target elements are
partially covered, \emph{extrapolation} could be used to extend a solution
outside the domain, but for totally uncovered elements there is no clear
correspondence to elements in the donor mesh.

@@

A \emph{B\'{e}zier curve} is a mapping from the unit interval
that is determined by a set of control points
\(\left\{\bm{p}_j\right\}_{j = 0}^n \subset \reals^d\).
For a parameter \(s \in \left[0, 1\right]\), there is a corresponding
point on the curve:
\begin{equation}
b(s) = \sum_{j = 0}^n \binom{n}{j} (1 - s)^{n - j} s^j \bm{p}_j \in
  \reals^d.
\end{equation}
This is a combination of the control points weighted by
each Bernstein basis function
\(B_{j, n}(s) = \binom{n}{j} (1 - s)^{n - j} s^j\).
Due to the binomial expansion
\(1 = (s + (1 - s))^n = \sum_{j = 0}^n B_{j, n}(s)\),
a Bernstein basis function is in
\(\left[0, 1\right]\) when \(s\) is as well. Due to this fact, the
curve must be contained in the convex hull of \rev{its} control points.

@@

A \emph{valid} B\'{e}zier triangle is one which is
diffeomorphic to \(\utri\), i.e. \(b(s, t)\) is bijective and has
an \rev{invertible Jacobian everywhere}. We must also have the orientation
preserved, i.e. the Jacobian must have positive determinant. For example, in
Figure~\ref{fig:inverted-element}, the image of \(\utri\) under
the map \(b(s, t) = \left[\begin{array}{c c} (1 - s - t)^2 + s^2 & s^2 + t^2
\end{array}\right]^T\) is not valid because the Jacobian is zero along
the curve \(s^2 - st - t^2 - s + t = 0\) (the dashed line). Elements that
are not valid are called \emph{inverted} because they have regions with
``negative area''. For the example, the image \(b\left(\utri\right)\)
leaves the boundary determined by the edge curves: \(b(r, 0)\),
\(b(1 - r, r)\) and \(b(0, 1 - r)\) when \(r \in \left[0, 1\right]\).
This region outside the boundary is traced twice, once with
a positive Jacobian and once with a negative Jacobian.
\begin{figure}
  \includegraphics{inverted_element.pdf}
  \centering
  \captionsetup{width=.75\linewidth}
  \caption{The B\'{e}zier triangle given by \(b(s, t) = \left[
    (1 - s - t)^2 + s^2 \; \; s^2 + t^2 \right]^T\) produces an
    inverted element. It traces the same region twice, once with
    a positive Jacobian (the middle column) and once with a negative
    Jacobian (the right column).}
  \label{fig:inverted-element}
\end{figure}

@@

Consider a donor mesh \(\mathcal{M}_D\) with shape function basis
\(\phi_D^{(j)}\) and a known field \(\bm{q}_D = \sum_j d_j \phi_D^{(j)}\)
\footnote{This is somewhat a simplification. \rev{In CG (continuous Galerkin, as
opposed to DG or discontinuous Galerkin)}, some coefficients will be paired
with multiple shape functions, such as the coefficient at a vertex node.}
and a
target mesh \(\mathcal{M}_T\) with shape function basis \(\phi_T^{(j)}\)
(Figure~\ref{fig:donor-target-pair}).
Each shape function \(\phi\) corresponds to a
given isoparametric curved element (see Section~\ref{sec:curved-elements})
\(\mathcal{T}\) in one of these meshes and has
\(\operatorname{supp}(\phi) = \mathcal{T}\).
Additionally, the shape functions are polynomial degree \(p\) (see
Section~\ref{subsec:shape-functions} for a discussion of shape functions),
but the degree of the donor mesh need not be the same as that of the
target mesh. We assume that both meshes cover the same domain \(\Omega \subset
\reals^2\), however we really only require the donor mesh to cover the
target mesh.

We seek the \(L_2\)-optimal interpolant \(\bm{q}_T = \sum_j t_j \phi_T^{(j)}\):
\begin{equation}
\left \lVert \bm{q}_T - \bm{q}_D \right \rVert_2 =
\min_{\bm{q} \in \mathcal{V}_T}
\left \lVert \bm{q} - \bm{q}_D \right \rVert_2
\end{equation}
where \(\mathcal{V}_T = \operatorname{Span}_j\left\{\phi_T^{(j)}\right\}\)
is the function space defined on the target mesh. Since this is optimal in
the \(L_2\) sense, by differentiating with respect to each \(t_j\) in
\(\bm{q}_T\) we find the weak form:
\begin{equation}
\int_{\Omega} \bm{q}_D \phi_T^{(j)} \, dV =
  \int_{\Omega} \bm{q}_T \phi_T^{(j)} \, dV, \qquad \text{for all } j.
\end{equation}
If the constant function \(1\) is contained in \(\mathcal{V}_T\),
conservation follows from the weak form and linearity of the integral
\begin{equation}
\int_{\Omega} \bm{q}_D \, dV =
  \int_{\Omega} \bm{q}_T \, dV.
\end{equation}
Expanding \(\bm{q}_D\) and \(\bm{q}_T\) with respect to their coefficients
\(\bm{d}\) and \(\bm{t}\), the weak form gives rise to a linear system
\begin{equation}\label{eq:weak-form-system}
M_T \bm{t} = M_{TD} \bm{d}.
\end{equation}
Here \(M_T\) is the mass matrix for
\(\mathcal{M}_T\) given by
\begin{equation}
\left(M_T\right)_{ij} = \int_{\Omega} \phi_T^{(i)} \phi_T^{(j)} \, dV.
\end{equation}
In the discontinuous Galerkin case, \(M_T\) is block diagonal with blocks
that correspond to each element, so \eqref{eq:weak-form-system} can be
solved locally on each element \(\mathcal{T}\) in the target mesh. By
construction, \(M_T\) is symmetric and sparse since \(\left(M_T\right)_{ij}\)
will be \(0\) unless \(\phi_T^{(i)}\) and \(\phi_T^{(j)}\) are supported
on the same element \(\mathcal{T}\). In the continuous case, \(M_T\) is
globally coupled since coefficients corresponding to boundary nodes interact
with multiple elements. The matrix \(M_{TD}\) is a ``mixed'' mass matrix
between the target and donor meshes:
\begin{equation}
\left(M_{TD}\right)_{ij} = \int_{\Omega} \phi_T^{(i)} \phi_D^{(j)} \, dV.
\end{equation}
Boundary conditions can be imposed on the system by fixing some coefficients,
but that is equivalent to removing some of the basis functions which may
\rev{in turn} make the projection non-conservative. This is because the removed
basis functions may have been used in \(1 = \sum_j u_j \phi_{T}^{(j)}\).

@@

For special cases, e.g. ALE methods, the initial brute-force search
can be reduced to \(\bigO{1}\). This could be enabled by tracking
the remeshing process so a correspondence already exists. If a full
mapping from donor to target mesh exists, the process of computing
\(M_T\) and \(M_{TD} \bm{d}\) can be fully parallelized across
elements of the target mesh, or even across shape functions
\(\phi_T^{(j)}\). \rev{Spatial binning techniques for mesh elements could also be
used to improve the initial brute-force search as in \cite{Weiss16_imr}.}

@@

Convergence results are shown in Figure~\ref{fig:composite-errors}
and confirm the expected orders. Since \(\zeta_1\) is a cubic polynomial,
one might expect the solution transfer on the cubic mesh to be \emph{exact}.
This expectation would hold in the superparametric case (i.e. where the
elements are straight sided but the shape functions are higher degree).
However, when the elements aren't straight \rev{sided} the isoparametric mapping
changes the underlying function space.

@@

\rev{In all of our experiments, we did not encounter any of these difficulties with our implementation. However, for fail-safe code, additional fallback strategies would be required, and we refer the reader to the extensive literature for further details on covering all possible special cases that might occur \cite{Sederberg1986,Sederberg1990,Kim1998}.}
